Bulk Semantic Annotation with an Partially Known Knowledge Base

### 2024-11-14 doc/cicm25-sysdesc/sysdesc.tex
more text

\documentclass{llncs}
\usepackage[show]{ed}
\usepackage{cleveref}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}

\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}

@@

In this paper, we address tools that help attack the coverage problem in symbolic/semantic
approaches in practice. The producer task we use as a case study is that of adaptive
learning assistants for MINT subjects, concretely the \ALeA system
\cite{BerBetChu:lssmkm23}, which uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
variant of {\LaTeX} that allows to embed semantic annotations -- for a knowledge
representation and generates learner-adaptive learning objects instrumented with
learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
\ednote{MK@MK: continue to describe the size and distribution, and
  authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
  \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
  releative sizes.}

@@

Concretely this is about converting the {\LaTeX} string\ednote{add numbers, start with 3
  so that corresponding lines with the next listing.}
\begin{lstlisting}[language={[LaTeX]TeX}]
\item \blue{Problem:} The state space has local minima, e.g. the board on the
  right has $h=1$ but every successor has $h>1$. 
\end{lstlisting}
into the (partially annotated item:
\begin{lstlisting}[language={[LaTeX]TeX},morekeywords={sn,importmodule}]
\importmodule[smglom/search]{mod?state-space}
[...]  
\item \blue{Problem:} The \sn{state space} has local minima, e.g. the
  board on the right has $h=1$ but every successor has $h>1$. 
\end{lstlisting}
Concretely, the author has to remember the module for the term ``state space'' from
\cref{fig:state-space}, the symbol name \lstinline|state space|, annotate it with the
\sTeX macro \lstinline|\sn{...}|, and add the module import in the first line (unless it
is already imported). In this case the symbol name happens to be the same as its English
verbalization, otherwise we would have to use
\lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm]{../img/state-space}}
  \caption{The definition of the term ``state space'' from the domain model}\label{fig:state-space}
\end{figure}

For this, the author has to be aware ``state space'' is a technical term, that the module
from \cref{fig:state-space} exists, know the symbol name and its URL, and manage
reduncancy of the imports. At ca 150.000 words in a semester's words of lecture notes and
ca. 10-15\% of technical terms a rather daunting task.

@@

The \srf system (see \cite{stextools:git} is a simple command-line tool that creates a
database of symbol-verbalization pairs\ednote{explain that first}, analyzes \sTeX source
files and then steps through all word occurrences in the document that share the stem with
a verbalization in the database. For each such word, the user is presented with an
annotation choice and interactions that allow to fine-tune the wordwise annotation
workflow.

As \srf addresses mostly power-users, the UI of \srf is command-line based, much like the
early \textsf{ispell} system, see\ednote{continue}

@@

In this system description we have presented the \srf tool, a command-line tool
\ednote{continue}

In an earlier attempt to support semantic annotation in IDEs, we tried named-entity
recognition (NER) for classification of ``likely annotation candidate words''
\cite{hutterer:msc23}, however this classification was not precise enough in the
distinction in ``technical terms'' and ordinary english noun phrases and named entities --
the relevant task for annotation, and so made it impractical, especially, since it could
not enter the actual annotations and import directives automatically. But maybe using the
NER-based approach, after the \srf annotation process might change the tradeoffs
involved. In any case, the overall workflow suggested by the NER-based approach --
especially after the verbalizations covered by the existing domain model\ednote{introduce
  in the definition above that the domain model is all definitions, not only in the
  smglom?} -- is geared towards adding modules and symbols -- the ones discovered by NER
-- to the domain model.

@@

% LocalWords:  tbw Aarne Ranta

>>>>

\documentclass{llncs}
\usepackage[show]{ed}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
%\usepackage{lststex} debug that the gray is not solid
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
\lstset{numberstyle=\tiny}
\lstset{language={[LaTeX]TeX}}
\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}
\usepackage[noabbrev]{cleveref}

@@

In this paper, we address tools that help attack the coverage problem in symbolic/semantic
approaches in practice. The producer task we use as a case study is that of adaptive
learning assistants for MINT subjects, concretely the \ALeA system
\cite{BerBetChu:lssmkm23}, which uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
variant of {\LaTeX} that allows to embed semantic annotations -- for a knowledge
representation and generates learner-adaptive learning objects instrumented with
learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
\ednote{MK@MK: continue to describe the size and distribution, and
  authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
  \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
  relative sizes.}

@@

Concretely this is about converting the {\LaTeX} string\ednote{add numbers, start with 3
  so that corresponding lines with the next listing.}
\begin{lstlisting}[numbers=left,firstnumber=3,
caption=The {\LaTeX} sources of \cref{fig:lo},label=lst:los]
\item \blue{Problem:} The state space has local minima, e.g. the board on the
  right has $h=1$ but every successor has $h>1$. 
\end{lstlisting}
into the (partially) annotated item:
\begin{lstlisting}[morekeywords={sn},numbers=left,
caption=Annotating ``state space'' in \cref{lst:los},label=lst:losa]
\importmodule[smglom/search]{mod?state-space}
[...]  
\item \blue{Problem:} The \sn{state space} has local minima, e.g. the
  board on the right has $h=1$ but every successor has $h>1$. 
\end{lstlisting}
Concretely, the author has to remember the module for the term ``state space'' from
\cref{fig:state-space}, the symbol name \lstinline|state space|, annotate it with the
\sTeX macro \lstinline|\sn{...}|, and add the module import in the first line (unless it
is already imported). In this case the symbol name happens to be the same as its English
verbalization, otherwise we would have to use
\lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.

The annotation is legitimized by a definition in the domain model module shown in
\cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces a
\textbf{verbalization} ``state space'' for the symbol \lstinline|state-space?state space|
introduced be \lstinline|\symdecl*| declaration in line 3.

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm]{../img/state-space}}
\begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
\begin{smodule}[title=State Space]{state-space}
[... some imports ...]
\symdecl*{state space}
\begin{sdefinition}
  [...] we call the \sn{graph} [...] the \definame{state space} of $\Pi$.
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{The definition of the term ``state space'' from the domain model}\label{fig:state-space}
\end{figure}

For the annotation \lstinline|\sn{state-space}|, the author has to be aware ``state
space'' is a technical term, that the module from \cref{fig:state-space} exists, know the
symbol name and its URL, and manage redundancy of the imports -- i.e. only adding the
\lstinline|\importmodule| directive if it is not already (recursively) implied and
possibly removing directives that become redundant by the new one. At ca 150.000 words in
a semester's words of lecture notes and ca. 10-15\% of technical terms a rather daunting
task.

@@

The \srf system (see \cite{stextools:git} is a simple command-line tool that creates a
database of symbol-verbalization pairs, analyzes \sTeX source files and then steps through
all word occurrences in the document that share the stem with a verbalization in the
database. For each such word, the user is presented with an annotation choice and
interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:srify}
shows a typical situation.\ednote{MK: I like the example less and less! We need an example
  that has at least one homostem (another verbalization that has the same stem; introduce
  the concept her and use it throughout) so that we can show the choice mechanism, and
  where the verbalization is different from the symbol name. Having a two-word compound is
  good though.}

\begin{figure}[ht]
  \fbox{\includegraphics[width=12cm]{../img/srify}}
  \caption{\srf in Action: Annotating the term ``state space''.}\label{fig:srify}
\end{figure}

Here the annotator can choose an annotation by typing the corresponding choice
number. Here, the module \lstinline|state-space| is already imported (see the little green
check-mark); otherwise \srf would offer the annotator the choice to import it -- if we are
in a module context -- or to add a \lstinline|\usemodule| directive and where (in the
local environment or top-level). Alternatively, typing \lstinline|s| skips this word,
\lstinline|S| skips and adds a comment to always skip this word, and -- if \srf first
suggests the term `state` \lstinline|n| extends it to the compound ``state space'' by
including the next word.

\srf is usually called on a set of files -- e.g. in a directory or math archive -- which
together form a \textbf{session}, which can be interrupted and resumed without having to
re-do all the \lstinline|s| skips. Alternatively, \srf can be called e.g. on the top-level
lecture notes; then the session consists of all included files. Other productivity
features include a focus mode that can be used to propagate a particular choice to the
rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
annotation This reduces the cognitive effort from choosing a symbol from a list to
deciding whether a particular choice is applicable in the current document context. Of
course the include management still needs to be done locally.

@@

In this system description we have presented the \srf tool, a command-line tool
\ednote{continue} \srf is open source\ednote{MK@FS: do we have an explicit license yet in
  the repository.}; the source and documentation are available from \cite{stextools:git}.

In an earlier attempt to support semantic annotation in IDEs, we tried named-entity
recognition (NER) for classification of ``likely annotation candidate words''
\cite{hutterer:msc23}, however this classification was not precise enough in the
distinction in ``technical terms'' and ordinary English noun phrases and named entities --
the relevant task for annotation, and so made it impractical, especially, since it could
not enter the actual annotations and import directives automatically. But maybe using the
NER-based approach, after the \srf annotation process might change the trade-offs
involved. In any case, the overall workflow suggested by the NER-based approach --
especially after the verbalizations covered by the existing domain model\ednote{introduce
  in the definition above that the domain model is all definitions, not only in the
  smglom?} -- is geared towards adding modules and symbols -- the ones discovered by NER
-- to the domain model.

@@

% LocalWords:  tbw Aarne Ranta homostem pre
